---
title: "Statistics"
output: html_document
date: "2025-12-30"
---
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(width=200)

```
# Statistical analyses

This document incudes the code for running all the statistical analyses of the mealworm and superworm gut microbiome data reported in this study. It includes the alpha-diversity, beta-diversity, and differential abundance analyses.


# Load required libraries

```{r Load libraries, results='hide', message=FALSE, warning=FALSE}

library(lme4)
library(phyloseq)
library(car)
library(emmeans)
library(ggplot2)
library(patchwork)
library(vegan)
library(dplyr)
library(tidyr)
library(kableExtra)
library(ANCOMBC)

```

# Analyses of the standard diet across species
## Prepare a phyloseq object for the standard diets for both insects

Use the MAG abundance data and the GTDB-Tk output data to generate a phyloseq object.

```{r Standard phyloseq object, results='hide', message=FALSE, warning=FALSE}

# Load the OTU table
otumat <- as.matrix(read.table("standard_MAG_abundances_depth.txt",
                        header = TRUE, row.names = 1))

# Load the taxonomy table
taxmat <- as.matrix(read.table("standard_MAG_taxonomy.txt",
                        header = TRUE, row.names = 1))

# Load the sample metadata
metadata <- read.table("standard_sample_metadata.txt",
                       header = TRUE, row.names = 1)

# Combine the data into a phyloseq object
OTU <- otu_table(otumat, taxa_are_rows = TRUE)
TAX <- tax_table(taxmat)
META <- sample_data(metadata)
ps <- phyloseq(OTU, TAX, META)

# Convert MAG abundances to percentages
ps.rel <- transform_sample_counts(ps, function(x) x / sum(x))

```

## Calculate alpha-diversity for the standard diets and perform statistical analysis

Calculate Shannon and Simpson's diversity metrics and test if the values are statistically different between samples. Include insect batch and replicate as random effects, with replicate nested within batch.

```{r Standard alpha diversity stats, message=FALSE, warning=FALSE}

# Calculate alpha-diversity metrics
richness <- estimate_richness(ps.rel, measures = c("Shannon", "Simpson"))

# Add columns with relevant metadata
richness$insect <- metadata$Insect
richness$replicate <- metadata$Replicate
richness$batch <- metadata$Insect_batch

# Create factors for treatment, ordering the treatments as desired
richness$treatment <- factor(richness$insect, 
                                 levels = c("Mealworm", "Superworm"))

# Perform statistical analysis on the Shannon data
model.shannon <- lmer(log(Shannon) ~ insect + (1 | replicate), data=richness)
summary(model.shannon)
model.shannon.aov <- Anova(model.shannon, type = 2)
model.shannon.aov
model.shannon.em <- emmeans(model.shannon, pairwise ~ insect,
                            infer = TRUE, type = "response")
model.shannon.em$emmeans
model.shannon.em$contrasts

# Perform statistical analysis on the Simpson data
model.simpson <- lmer(log(Simpson) ~ insect + (1 | replicate), data=richness)
summary(model.simpson)
model.simpson.aov <- Anova(model.simpson, type = 2)
model.simpson.aov
model.simpson.em <- emmeans(model.simpson, pairwise ~ insect,
                            infer = TRUE, type = "response")
model.simpson.em$emmeans
model.simpson.em$contrasts

```

## Plot the alpha-diversity metrics for the standard diets

Generate plots showing the Shannon and Simpson's diversity metrics

```{r Standard alpha diversity plots, results='hide', message=FALSE, warning=FALSE}

# Collect the emmeans for Shannon data
emms.shannon <- as.data.frame(model.shannon.em$emmeans)
emms.shannon$metric <- 'Shannon'

# Collect the emmeans for teh Simpson's data
emms.simpson <- as.data.frame(model.simpson.em$emmeans)
emms.simpson$metric <- 'Simpson'

# Plot the Shannon data
p.shannon <- ggplot() +
  geom_bar(data = emms.shannon,
           aes(x = insect, y = response, fill = insect),
           stat = "identity",
           colour = "black") +
  geom_errorbar(data = emms.shannon,
                aes(x = insect, ymin = response-SE, ymax = response+SE),
                stat = "identity", width = 0.5, size = 0.5) +
  geom_jitter(data = richness,
              aes(x = insect, y = Shannon),
              width = 0.15, size = 1.25) +
  facet_wrap(~metric, scales="free_y") +
  theme_bw() +
  theme(text = element_text(size = 17), legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(face="bold"),
        axis.text.x = element_text(color="black"),
        axis.text.y = element_text(color="black")) +
  scale_y_continuous(lim=c(0,3.5), breaks=c(0.0,0.7,1.4,2.1,2.8,3.5)) + 
  xlab("") + ylab("")

# Plot the Simpson's data
p.simpson <- ggplot() +
  geom_bar(data = emms.simpson,
           aes(x = insect, y = response, fill = insect),
           stat = "identity",
           colour = "black") +
  geom_errorbar(data = emms.simpson,
                aes(x = insect, ymin = response-SE, ymax = response+SE),
                stat = "identity", width = 0.5, size = 0.5) +
  geom_jitter(data = richness,
              aes(x = insect, y = Simpson),
              width = 0.15, size = 1.25) +
  facet_wrap(~metric, scales="free_y") +
  theme_bw() +
  theme(text = element_text(size = 17), legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(face="bold"),
        axis.text.x = element_text(color="black"),
        axis.text.y = element_text(color="black")) +
  scale_y_continuous(lim=c(0,1), breaks=c(0.0,0.2,0.4,0.6,0.8,1.0)) + 
  xlab("") + ylab("")

# Combine the plots as two panels in one figure
p.shannon + p.simpson

# Export a PDF of the figure
pdf('Standard.alpha.pdf', width=10, height=5)
p.shannon + p.simpson
dev.off()

```

## Calculate beta-diversity and make a PCoA plot at the species level to compare insect species

```{r Standard beta diversity species, message=FALSE, warning=FALSE}

# Summarize at the species level, removing taxa not classified at the species level
ps.rel.sp <- tax_glom(ps.rel, "Species", NArm = TRUE)

# Calculate the bray curtis distances
dist_bc <- phyloseq::distance(ps.rel.sp, method = "bray")

# Extract sample data
meta <- data.frame(sample_data(ps.rel.sp))

# Perform a permutation test to test if the data meet the assumptions of PERMANOVA
beta.bray <- betadisper(dist_bc, meta$Insect, type = "centroid", bias.adjust = TRUE)
permdisp <- permutest(beta.bray, pairwise = TRUE, permutations = 10000)
permdisp

# Run the PERMANOVA test
adonis.bray <- adonis2(dist_bc ~ Insect, data = meta, permutations = 10000, by = "terms")
as.data.frame(adonis.bray)

# Create a PCoA plot
pso <- ordinate(ps.rel.sp, "PCoA", "bray")
var_explained <- pso$values$Relative_eig * 100
ordplot5 <- plot_ordination(ps.rel.sp, pso, type="samples", color="Insect", axes = c(1,2))
p.beta.sp <- ordplot5 +  theme_bw() + theme(text = element_text(size = 17), aspect.ratio = 1) + 
  geom_point(size = 4) +
  geom_point(size = 4, shape=21, stroke=0.5, color="black") +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = paste0(" PC1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" PC2 [", round(var_explained[2], 1), "%]")) +
  scale_x_continuous(lim=c(-0.50,0.50), breaks=c(-0.50,-0.25,0.00,0.25,0.50),
                     labels = function(x) sprintf("%.2f", x)) +
  scale_y_continuous(lim=c(-0.25,0.75), breaks=c(-0.25,0.00,0.25,0.50,0.75),
                     labels = function(x) sprintf("%.2f", x))

```

## Calculate beta-diversity and make a PCoA plot at the genus level to compare insect species

```{r Standard beta diversity genus, message=FALSE, warning=FALSE}

# Summarize at the species level, removing taxa not classified at the species level
ps.rel.genus <- tax_glom(ps.rel, "Genus", NArm = TRUE)

# Calculate the bray curtis distances
dist_bc <- phyloseq::distance(ps.rel.genus, method = "bray")

# Extract sample data
meta <- data.frame(sample_data(ps.rel.genus))

# Perform a permutation test to test if the data meet the assumptions of PERMANOVA
beta.bray <- betadisper(dist_bc, meta$Insect, type = "centroid", bias.adjust = TRUE)
permdisp <- permutest(beta.bray, pairwise = TRUE, permutations = 10000)
permdisp

# Run the PERMANOVA test
adonis.bray <- adonis2(dist_bc ~ Insect, data = meta, permutations = 10000, by = "terms")
as.data.frame(adonis.bray)

# Create a PCoA plot
pso <- ordinate(ps.rel.genus, "PCoA", "bray")
var_explained <- pso$values$Relative_eig * 100
ordplot5 <- plot_ordination(ps.rel.genus, pso, type="samples", color="Insect", axes = c(1,2))
p.beta.genus <- ordplot5 +  theme_bw() + theme(text = element_text(size = 17), aspect.ratio = 1) + 
  geom_point(size = 4) +
  geom_point(size = 4, shape=21, stroke=0.5, color="black") +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = paste0(" PC1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" PC2 [", round(var_explained[2], 1), "%]")) +
  scale_x_continuous(lim=c(-0.60,0.60), breaks=c(-0.60,-0.30,0.00,0.30,0.60),
                     labels = function(x) sprintf("%.2f", x)) +
  scale_y_continuous(lim=c(-0.50,0.50), breaks=c(-0.50,-0.25,0.00,0.25,0.50),
                     labels = function(x) sprintf("%.2f", x))

```

## Calculate beta-diversity and make a PCoA plot at the family level to compare insect species

```{r Standard beta diversity family, message=FALSE, warning=FALSE}

# Summarize at the species level, removing taxa not classified at the species level
ps.rel.fam <- tax_glom(ps.rel, "Family", NArm = TRUE)

# Calculate the bray curtis distances
dist_bc <- phyloseq::distance(ps.rel.fam, method = "bray")

# Extract sample data
meta <- data.frame(sample_data(ps.rel.fam))

# Perform a permutation test to test if the data meet the assumptions of PERMANOVA
beta.bray <- betadisper(dist_bc, meta$Insect, type = "centroid", bias.adjust = TRUE)
permdisp <- permutest(beta.bray, pairwise = TRUE, permutations = 10000)
permdisp

# Run the PERMANOVA test
adonis.bray <- adonis2(dist_bc ~ Insect, data = meta, permutations = 10000, by = "terms")
as.data.frame(adonis.bray)

# Create a PCoA plot
pso <- ordinate(ps.rel.fam, "PCoA", "bray")
var_explained <- pso$values$Relative_eig * 100
ordplot5 <- plot_ordination(ps.rel.fam, pso, type="samples", color="Insect", axes = c(1,2))
p.beta.fam <- ordplot5 +  theme_bw() + theme(text = element_text(size = 17), aspect.ratio = 1) + 
  geom_point(size = 4) +
  geom_point(size = 4, shape=21, stroke=0.5, color="black") +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = paste0(" PC1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" PC2 [", round(var_explained[2], 1), "%]")) +
  scale_x_continuous(lim=c(-0.60,0.60), breaks=c(-0.60,-0.30,0.00,0.30,0.60),
                     labels = function(x) sprintf("%.2f", x)) +
  scale_y_continuous(lim=c(-0.30,0.50), breaks=c(-0.30,-0.10,0.10,0.30,0.50), 
                     labels = function(x) sprintf("%.2f", x))

```

## Calculate beta-diversity and make a PCoA plot at the order level to compare insect species

```{r Standard beta diversity order, message=FALSE, warning=FALSE}

# Summarize at the species level, removing taxa not classified at the species level
ps.rel.order <- tax_glom(ps.rel, "Order", NArm = TRUE)

# Calculate the bray curtis distances
dist_bc <- phyloseq::distance(ps.rel.order, method = "bray")

# Extract sample data
meta <- data.frame(sample_data(ps.rel.order))

# Perform a permutation test to test if the data meet the assumptions of PERMANOVA
beta.bray <- betadisper(dist_bc, meta$Insect, type = "centroid", bias.adjust = TRUE)
permdisp <- permutest(beta.bray, pairwise = TRUE, permutations = 10000)
permdisp

# Run the PERMANOVA test
adonis.bray <- adonis2(dist_bc ~ Insect, data = meta, permutations = 10000, by = "terms")
as.data.frame(adonis.bray)

# Create a PCoA plot
pso <- ordinate(ps.rel.order, "PCoA", "bray")
var_explained <- pso$values$Relative_eig * 100
ordplot5 <- plot_ordination(ps.rel.order, pso, type="samples", color="Insect", axes = c(1,2))
p.beta.order <- ordplot5 +  theme_bw() + theme(text = element_text(size = 17), aspect.ratio = 1) + 
  geom_point(size = 4) +
  geom_point(size = 4, shape=21, stroke=0.5, color="black") +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = paste0(" PC1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" PC2 [", round(var_explained[2], 1), "%]")) +
  scale_x_continuous(lim=c(-0.50,0.50), breaks=c(-0.50,-0.25,0.00,0.25,0.50),
                     labels = function(x) sprintf("%.2f", x)) +
  scale_y_continuous(lim=c(-0.40,0.40), breaks=c(-0.40,-0.20,0.00,0.20,0.40), 
                     labels = function(x) sprintf("%.2f", x))

```

## Calculate beta-diversity and make a PCoA plot at the class level to compare insect species

```{r Standard beta diversity class, message=FALSE, warning=FALSE}

# Summarize at the species level, removing taxa not classified at the species level
ps.rel.class <- tax_glom(ps.rel, "Class", NArm = TRUE)

# Calculate the bray curtis distances
dist_bc <- phyloseq::distance(ps.rel.class, method = "bray")

# Extract sample data
meta <- data.frame(sample_data(ps.rel.class))

# Perform a permutation test to test if the data meet the assumptions of PERMANOVA
beta.bray <- betadisper(dist_bc, meta$Insect, type = "centroid", bias.adjust = TRUE)
permdisp <- permutest(beta.bray, pairwise = TRUE, permutations = 10000)
permdisp

# Run the PERMANOVA test
adonis.bray <- adonis2(dist_bc ~ Insect, data = meta, permutations = 10000, by = "terms")
as.data.frame(adonis.bray)

# Create a PCoA plot
pso <- ordinate(ps.rel.class, "PCoA", "bray")
var_explained <- pso$values$Relative_eig * 100
ordplot5 <- plot_ordination(ps.rel.class, pso, type="samples", color="Insect", axes = c(1,2))
p.beta.class <- ordplot5 +  theme_bw() + theme(text = element_text(size = 17), aspect.ratio = 1) + 
  geom_point(size = 4) +
  geom_point(size = 4, shape=21, stroke=0.5, color="black") +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = paste0(" PC1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" PC2 [", round(var_explained[2], 1), "%]")) +
  scale_x_continuous(lim=c(-0.50,0.50), breaks=c(-0.50,-0.25,0.00,0.25,0.50),
                     labels = function(x) sprintf("%.2f", x)) +
  scale_y_continuous(lim=c(-0.20,0.20), breaks=c(-0.10,-0.10,0.00,0.10,0.20), 
                     labels = function(x) sprintf("%.2f", x))

```

## Calculate beta-diversity and make a PCoA plot at the phylum level to compare insect species

```{r Standard beta diversity phylum, message=FALSE, warning=FALSE}

# Summarize at the species level, removing taxa not classified at the species level
ps.rel.phylum <- tax_glom(ps.rel, "Phylum", NArm = TRUE)

# Calculate the bray curtis distances
dist_bc <- phyloseq::distance(ps.rel.phylum, method = "bray")

# Extract sample data
meta <- data.frame(sample_data(ps.rel.phylum))

# Perform a permutation test to test if the data meet the assumptions of PERMANOVA
beta.bray <- betadisper(dist_bc, meta$Insect, type = "centroid", bias.adjust = TRUE)
permdisp <- permutest(beta.bray, pairwise = TRUE, permutations = 10000)
permdisp

# Run the PERMANOVA test
adonis.bray <- adonis2(dist_bc ~ Insect, data = meta, permutations = 10000, by = "terms")
as.data.frame(adonis.bray)

# Create a PCoA plot
pso <- ordinate(ps.rel.phylum, "PCoA", "bray")
var_explained <- pso$values$Relative_eig * 100
ordplot5 <- plot_ordination(ps.rel.phylum, pso, type="samples", color="Insect", axes = c(1,2))
p.beta.phylum <- ordplot5 +  theme_bw() + theme(text = element_text(size = 17), aspect.ratio = 1) + 
  geom_point(size = 4) +
  geom_point(size = 4, shape=21, stroke=0.5, color="black") +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = paste0(" PC1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" PC2 [", round(var_explained[2], 1), "%]")) +
  scale_x_continuous(lim=c(-0.50,0.50), breaks=c(-0.50,-0.25,0.00,0.25,0.50),
                     labels = function(x) sprintf("%.2f", x)) +
  scale_y_continuous(lim=c(-0.20,0.20), breaks=c(-0.20,-0.10,0.00,0.10,0.20), 
                     labels = function(x) sprintf("%.2f", x))

```

## Make a figure of all the PCoA plots at different taxonomic levels

```{r Standard beta diversity plot, results=FALSE, message=FALSE, warning=FALSE}

p.beta.sp + p.beta.genus + p.beta.fam + p.beta.order + p.beta.class + p.beta.phylum

pdf('Standard.beta.pdf', width=9, height=5)
p.beta.sp + p.beta.genus + p.beta.fam + p.beta.order + p.beta.class + p.beta.phylum
dev.off()


```

## Compare taxon abundance in the standard data

Summarize the most abundant taxa at the phylum level in the mealworms and superworms in the standard diets.

```{r Standard phylum abundance, message=FALSE, warning=FALSE}

# Agglomerate data at phylum level 
ps.rel.phylum <- tax_glom(ps.rel, "Phylum", NArm = FALSE)

# Create a dataframe from phyloseq object
ps.rel.phylum.df <- psmelt(ps.rel.phylum)

# Get summary statistics
df <- ps.rel.phylum.df %>%
  group_by(Insect, Phylum) %>%
  summarize(mean = mean(Abundance, na.rm = TRUE),
            median = median(Abundance, na.rm = TRUE),
            stdev = sd(Abundance, na.rm = TRUE),
            counts = n(),
            se = stdev / sqrt(counts))

# Create a table rounded to 1 decimal place
df.w <- df %>%
  mutate(stat = paste0(round(100 * mean, 1), " +/- ", round(100 * stdev, 1))) %>%
  select(Insect, Phylum, stat) %>%
  pivot_wider(
    names_from = Insect,
    values_from = stat
  )
kable(df.w, digits = 1, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

# Create a table rounded to 6 decimal place
df.w <- df %>%
  mutate(stat = paste0(round(100 * mean, 6), " +/- ", round(100 * stdev, 6))) %>%
  select(Insect, Phylum, stat) %>%
  pivot_wider(
    names_from = Insect,
    values_from = stat
  )
kable(df.w, digits = 6, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

```

Summarize the most abundant taxa at the order level in the mealworms and superworms in the standard diets for the 10 most abundant orders.

```{r Standard order abundance, message=FALSE, warning=FALSE}

# Agglomerate data at order level 
ps.rel.order <- tax_glom(ps.rel, "Order", NArm = FALSE)

# Identify the most abundant orders
top.taxa = names(sort(taxa_sums(ps.rel.order), TRUE)[1:10])
top.taxa.table = cbind(tax_table(ps.rel.order), topTaxa = NA)
top.taxa.table[top.taxa, "topTaxa"] <- as(tax_table(ps.rel.order)[top.taxa, "Order"], 
                                          "character")
tax_table(ps.rel.order) <- tax_table(top.taxa.table)

# Create a dataframe from phyloseq object
ps.rel.order.df <- psmelt(ps.rel.order)
ps.rel.order.df$Order <- as.character(ps.rel.order.df$Order) #convert to character
ps.rel.order.df.noNA <- na.omit(ps.rel.order.df)
ps.rel.order.df <- ps.rel.order.df.noNA

# Get summary statistics
df <- ps.rel.order.df %>%
  group_by(Insect, Order) %>%
  summarize(mean = mean(Abundance, na.rm = TRUE),
            median = median(Abundance, na.rm = TRUE),
            stdev = sd(Abundance, na.rm = TRUE),
            counts = n(),
            se = stdev / sqrt(counts))

# Create a table rounded to 1 decimal place
df.w <- df %>%
  mutate(stat = paste0(round(100 * mean, 1), " +/- ", round(100 * stdev, 1))) %>%
  select(Insect, Order, stat) %>%
  pivot_wider(
    names_from = Insect,
    values_from = stat
  )
kable(df.w, digits = 1, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

# Create a table rounded to 6 decimal place
df.w <- df %>%
  mutate(stat = paste0(round(100 * mean, 6), " +/- ", round(100 * stdev, 6))) %>%
  select(Insect, Order, stat) %>%
  pivot_wider(
    names_from = Insect,
    values_from = stat
  )
kable(df.w, digits = 6, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

```

Summarize the most abundant taxa at the genus level in the mealworms and superworms in the standard diets for the 40 most abundant genera

```{r Standard genus abundance, message=FALSE, warning=FALSE}

# Agglomerate data at genus level 
ps.rel.genus <- tax_glom(ps.rel, "Genus", NArm = FALSE)

# Identify the most abundant orders
top.taxa = names(sort(taxa_sums(ps.rel.genus), TRUE)[1:40])
top.taxa.table = cbind(tax_table(ps.rel.genus), topTaxa = NA)
top.taxa.table[top.taxa, "topTaxa"] <- as(tax_table(ps.rel.genus)[top.taxa, "Genus"], 
                                          "character")
tax_table(ps.rel.genus) <- tax_table(top.taxa.table)

# Create a dataframe from phyloseq object
ps.rel.genus.df <- psmelt(ps.rel.genus)
ps.rel.genus.df$Genus <- as.character(ps.rel.genus.df$Genus) #convert to character
ps.rel.genus.df.noNA <- na.omit(ps.rel.genus.df)
ps.rel.genus.df <- ps.rel.genus.df.noNA

# Get summary statistics
df <- ps.rel.genus.df %>%
  group_by(Insect, Genus) %>%
  summarize(mean = mean(Abundance, na.rm = TRUE),
            median = median(Abundance, na.rm = TRUE),
            stdev = sd(Abundance, na.rm = TRUE),
            counts = n(),
            se = stdev / sqrt(counts))

# Create a table rounded to 1 decimal place
df.w <- df %>%
  mutate(stat = paste0(round(100 * mean, 1), " +/- ", round(100 * stdev, 1))) %>%
  select(Insect, Genus, stat) %>%
  pivot_wider(
    names_from = Insect,
    values_from = stat
  )
kable(df.w, digits = 1, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

# Create a table rounded to 6 decimal place
df.w <- df %>%
  mutate(stat = paste0(round(100 * mean, 6), " +/- ", round(100 * stdev, 6))) %>%
  select(Insect, Genus, stat) %>%
  pivot_wider(
    names_from = Insect,
    values_from = stat
  )
kable(df.w, digits = 6, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

```

## Clean the workspace to switch to mealworm data

```{r Clear standard data, results='hide', message=FALSE, warning=FALSE}

rm(list = ls())

```


# Analyse of the mealworm data across treatments
## Prepare a phyloseq object for the mealworm data

Use the MAG abundance data and the GTDB-Tk output data to generate a phyloseq object.

```{r Mealworm phyloseq object, results='hide', message=FALSE, warning=FALSE}

# Load the OTU table
otumat <- as.matrix(read.table("mealworm_MAG_abundances_depth.txt",
                        header = TRUE, row.names = 1))

# Load the taxonomy table
taxmat <- as.matrix(read.table("mealworm_MAG_taxonomy.txt",
                        header = TRUE, row.names = 1))

# Load the sample metadata
metadata <- read.table("mealworm_sample_metadata.txt",
                       header = TRUE, row.names = 1)

# Combine the data into a phyloseq object
OTU <- otu_table(otumat, taxa_are_rows = TRUE)
TAX <- tax_table(taxmat)
META <- sample_data(metadata)
ps <- phyloseq(OTU, TAX, META)

# Convert MAG abundances to percentages
ps.rel <- transform_sample_counts(ps, function(x) x / sum(x))

```

## Calculate alpha-diversity for the mealworm data and perform statistical analysis

Calculate Shannon and Simpson's diversity metrics and test if the values are statistically different between samples. Include insect batch and replicate as random effects, with replicate nested within batch.

```{r Mealworm alpha diversity stats, message=FALSE, warning=FALSE}


# Calculate alpha-diversity metrics
richness <- estimate_richness(ps.rel, measures = c("Shannon", "Simpson"))

# Add columns with relevant metadata
richness$treatment <- metadata$Treatment
richness$replicate <- metadata$Replicate
richness$batch <- metadata$Insect_batch

# Create factors for treatment, ordering the treatments as desired
richness$treatment <- factor(richness$treatment, 
                                 levels = c("Standard", "Polyethylene", "Polystyrene", "Starvation"))

# Perform statistical analysis on the Shannon data
model.shannon <- lmer(log(Shannon) ~ treatment + (1 | batch/replicate), data=richness)
summary(model.shannon)
model.shannon.aov <- Anova(model.shannon, type = 2)
model.shannon.em <- emmeans(model.shannon, pairwise ~ treatment,
                            infer = TRUE, type = "response")
model.shannon.em$emmeans
model.shannon.em$contrasts

# Perform statistical analysis on the Simpson data
model.simpson <- lmer(log(Simpson) ~ treatment + (1 | batch/replicate), data=richness)
summary(model.simpson)
model.simpson.aov <- Anova(model.simpson, type = 2)
model.simpson.em <- emmeans(model.simpson, pairwise ~ treatment,
                            infer = TRUE, type = "response")
model.simpson.em$emmeans
model.simpson.em$contrasts

```

## Plot the alpha-diversity metrics for the mealworm samples

Generate plots showing the Shannon and Simpson's diversity metrics

```{r Mealworm alpha diversity plots, results='hide', message=FALSE, warning=FALSE}

# Collect the emmeans for Shannon data
emms.shannon <- as.data.frame(model.shannon.em$emmeans)
emms.shannon$metric <- 'Shannon'

# Collect the emmeans for teh Simpson's data
emms.simpson <- as.data.frame(model.simpson.em$emmeans)
emms.simpson$metric <- 'Simpson'

# Plot the Shannon data
p.shannon <- ggplot() +
  geom_bar(data = emms.shannon,
           aes(x = treatment, y = response, fill = treatment),
           stat = "identity",
           colour = "black") +
  geom_errorbar(data = emms.shannon,
                aes(x = treatment, ymin = response-SE, ymax = response+SE),
                stat = "identity", width = 0.5, size = 0.5) +
  geom_jitter(data = richness,
              aes(x = treatment, y = Shannon),
              width = 0.15, size = 1.25) +
  facet_wrap(~metric, scales="free_y") +
  theme_bw() +
  theme(text = element_text(size = 17), legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(face="bold"),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1, color="black"),
        axis.text.y = element_text(color="black")) +
  scale_y_continuous(lim=c(0,3), breaks=c(0.0,0.6,1.2,1.8,2.4,3.0)) + 
  xlab("") + ylab("")

# Plot the Simpson's data
p.simpson <- ggplot() +
  geom_bar(data = emms.simpson,
           aes(x = treatment, y = response, fill = treatment),
           stat = "identity",
           colour = "black") +
  geom_errorbar(data = emms.simpson,
                aes(x = treatment, ymin = response-SE, ymax = response+SE),
                stat = "identity", width = 0.5, size = 0.5) +
  geom_jitter(data = richness,
              aes(x = treatment, y = Simpson),
              width = 0.15, size = 1.25) +
  facet_wrap(~metric, scales="free_y") +
  theme_bw() +
  theme(text = element_text(size = 17), legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(face="bold"),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1, color="black"),
        axis.text.y = element_text(color="black")) +
  scale_y_continuous(lim=c(0,1), breaks=c(0.0,0.2,0.4,0.6,0.8,1.0)) + 
  xlab("") + ylab("")

# Combine the plots as two panels in one figure
p.shannon + p.simpson

# Export a PDF of the figure
pdf('Mealworm.alpha.pdf', width=10, height=5)
p.shannon + p.simpson
dev.off()

```

## Calculate and plot beta-diversity for the mealworm data and perform statistical analysis

```{r Mealworm beta diversity permdisp, message=FALSE, warning=FALSE}

# Calculate the bray curtis distances
dist_bc <- phyloseq::distance(ps.rel, method = "bray")

# Extract sample data
meta <- data.frame(sample_data(ps.rel))

# Perform a permutation test to test if the data meet the assumptions of PERMANOVA
beta.bray <- betadisper(dist_bc, meta$Treatment, type = "centroid", bias.adjust = TRUE)
permdisp <- permutest(beta.bray, pairwise = TRUE, permutations = 10000)
permdisp

```

The dispersion is significantly different between samples, which means the results of PERMANOVA are not reliable. Nevertheless, run the PERMANOVA (accounting for treatment and batch) and pair it with a PCoA to visualize the data.

```{r Mealworm beta diversity permanova, message=FALSE, warning=FALSE}

# Run the PERMANOVA test
adonis.bray <- adonis2(dist_bc ~ Treatment + Insect_batch, data = meta, permutations = 10000, by = "terms")
as.data.frame(adonis.bray)

```

```{r Mealworm beta diversity figure, results='hide', message=FALSE, warning=FALSE}

# Create a PCoA plot
sample_data(ps.rel)$Treatment <- factor(sample_data(ps.rel)$Treatment, 
                                        levels = c("Standard", "Polyethylene", "Polystyrene", "Starvation"))
pso <- ordinate(ps.rel, "PCoA", "bray")
var_explained <- pso$values$Relative_eig * 100
ordplot5 <- plot_ordination(ps.rel, pso, type="samples", 
                            color="Treatment", shape="Insect_batch",
                            axes = c(1,2))
p.beta.pcoa <- ordplot5 +  theme_bw() + theme(text = element_text(size = 17), aspect.ratio = 1) + 
  geom_point(size = 5) +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right") +
  labs(x = paste0(" PC1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" PC2 [", round(var_explained[2], 1), "%]"))
p.beta.pcoa

# Save the figure
pdf('Mealworm.beta.pcoa.pdf', width=10, height=5)
p.beta.pcoa
dev.off()

```

Now run a capscale analysis since it is less sensitive to differences in dispersion and thus should be more appropriate than PERMANOVA in this case.

```{r Mealworm beta diversity capscale, message=FALSE, warning=FALSE}

# Set the formula for the model
class(fo <- dist_bc ~ meta$Treatment + meta$Insect_batch)

# Run the capscale analysis followed by an ANOVA
pso_cap <- ordinate(ps.rel, "CAP", "bray", formula=fo)
anova(pso_cap, by = "terms", type = 2)

```

```{r Mealworm beta diversity capscale figure, results='hide', message=FALSE, warning=FALSE}

# Get the relative eigenvalues
all_eig <- eigenvals(pso_cap)
rel_eig <- all_eig / sum(all_eig)
var_explained <- round(rel_eig * 100, 1)

# Plot the capscale results
ordplot5 <- plot_ordination(ps.rel, pso_cap, type="samples", color="Treatment",
                            shape="Insect_batch", axes = c(1,2))
p.beta.cap <- ordplot5 +  theme_bw() + theme(text = element_text(size = 17), aspect.ratio = 1) + 
  geom_point(size = 5) +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right") +
  labs(x = paste0(" CAP 1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" CAP 2 [", round(var_explained[2], 1), "%]"))
p.beta.cap

# Save the figure
pdf('Mealworm.beta.cap.pdf', width=10, height=5)
p.beta.pcoa
dev.off()

```

To mimic what will be done for the superworm data, see what the PCoA and CAP plots look like if summarized at the genus level

```{r Mealworm beta diversity genus, message=FALSE, warning=FALSE}

# Summarize at the genus level, removing taxa not classified at the genus level
ps.rel.genus <- tax_glom(ps.rel, "Genus", NArm = TRUE)

# Calculate the bray curtis distances
dist_bc <- phyloseq::distance(ps.rel.genus, method = "bray")

# Extract sample data
meta <- data.frame(sample_data(ps.rel.genus))

# Perform a permutation test to test if the data meet the assumptions of PERMANOVA
beta.bray <- betadisper(dist_bc, meta$Treatment, type = "centroid", bias.adjust = TRUE)
permdisp <- permutest(beta.bray, pairwise = TRUE, permutations = 10000)
permdisp

# Run the PERMANOVA test
adonis.bray <- adonis2(dist_bc ~ Treatment + Insect_batch, data = meta, permutations = 10000, by = "terms")
as.data.frame(adonis.bray)

# Create a PCoA plot
pso <- ordinate(ps.rel.genus, "PCoA", "bray")
var_explained <- pso$values$Relative_eig * 100
ordplot5 <- plot_ordination(ps.rel.genus, pso, type="samples", 
                            color="Treatment", shape="Insect_batch", 
                            axes = c(1,2))
p.beta.pcoa.genus <- ordplot5 + theme_bw() + 
  theme(text = element_text(size = 17), aspect.ratio = 1) +
  geom_point(size = 5) +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = paste0(" PC1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" PC2 [", round(var_explained[2], 1), "%]"))

# Set the formula for the model for the capscale analysis
class(fo <- dist_bc ~ meta$Treatment + meta$Insect_batch)

# Run the capscale analysis followed by an ANOVA
pso_cap <- ordinate(ps.rel.genus, "CAP", "bray", formula=fo)
anova(pso_cap, by = "terms", type = 2)

# Get the relative eigenvalues
all_eig <- eigenvals(pso_cap)
rel_eig <- all_eig / sum(all_eig)
var_explained <- round(rel_eig * 100, 1)

# Plot the capscale results
ordplot5 <- plot_ordination(ps.rel.genus, pso_cap, type="samples", color="Treatment",
                            shape="Insect_batch", axes = c(1,2))
p.beta.cap.genus <- ordplot5 + theme_bw() + 
  theme(text = element_text(size = 17), aspect.ratio = 1) +
  geom_point(size = 5) +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = paste0(" CAP 1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" CAP 2 [", round(var_explained[2], 1), "%]"))

```

```{r Mealworm beta diversity genus figure, results='hide', message=FALSE, warning=FALSE}

# Display both graphs
p.beta.pcoa.genus + p.beta.cap.genus

# Save the figure
pdf('Mealworm.beta.genus.pdf', width=10, height=5)
p.beta.pcoa.genus + p.beta.cap.genus
dev.off()

```

## Compare taxon abundance in the mealworm data

Summarize the most abundant taxa at the phylum level in the mealworms across diets.

```{r Mealworm phylum abundance, message=FALSE, warning=FALSE}

# Agglomerate data at phylum level 
ps.rel.phylum <- tax_glom(ps.rel, "Phylum", NArm = FALSE)

# Create a dataframe from phyloseq object
ps.rel.phylum.df <- psmelt(ps.rel.phylum)

# Get summary statistics
df <- ps.rel.phylum.df %>%
  group_by(Treatment, Phylum) %>%
  summarize(mean = mean(Abundance, na.rm = TRUE),
            median = median(Abundance, na.rm = TRUE),
            stdev = sd(Abundance, na.rm = TRUE),
            counts = n(),
            se = stdev / sqrt(counts))

# Create a table rounded to 1 decimal place
df.w <- df %>%
  mutate(stat = paste0(round(100 * mean, 1), " +/- ", round(100 * stdev, 1))) %>%
  select(Treatment, Phylum, stat) %>%
  pivot_wider(
    names_from = Treatment,
    values_from = stat
  )
kable(df.w, digits = 1, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

# Create a table rounded to 6 decimal place
df.w <- df %>%
  mutate(stat = paste0(round(100 * mean, 6), " +/- ", round(100 * stdev, 6))) %>%
  select(Treatment, Phylum, stat) %>%
  pivot_wider(
    names_from = Treatment,
    values_from = stat
  )
kable(df.w, digits = 6, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

# Get overall summary statistics not divided by diet
df <- ps.rel.phylum.df %>%
  group_by(Phylum) %>%
  summarize(mean = 100 * mean(Abundance, na.rm = TRUE),
            stdev = 100 * sd(Abundance, na.rm = TRUE),
            counts = n(),
            se = stdev / sqrt(counts))
kable(df, digits = 6, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

```

## Differential abundance analysis of taxa in the mealworm data

Test whether any of the MAGs are enriched in the PE or PS diets relative to both the starvation and standard diets.

```{r Mealworm MAG differential abundance, message=FALSE, warning=FALSE}

# Set abundance filter function, filtering based on an average relative abundance ≥ 0.01%
abundance_filter_function <- function(x) {
  mean(x) >= 0.0001
}

# Filter the MAG data to a minimum average relative abundance of 0.01%
ps.rel.filtered <- ps

# Set the factors to get standard as the reference
sample_data(ps.rel.filtered)$Treatment <- factor(sample_data(ps.rel.filtered)$Treatment, 
                                                 levels = c("Standard", "Polyethylene",
                                                            "Polystyrene","Starvation"))

# Run ANCOMBC2 with standard as teh reference
ancombc.standard <- ancombc2(data = ps.rel.filtered,
                             fix_formula = "Treatment", 
                             rand_formula = "(1 | Insect_batch/Replicate)",
                             p_adj_method = "BH",
                             alpha = 0.025,
                             global = FALSE)
ancombc.standard.res <- ancombc.standard$res

# Set the factors to get standard as the reference
sample_data(ps.rel.filtered)$Treatment <- factor(sample_data(ps.rel.filtered)$Treatment, 
                                                 levels = c("Starvation", "Polyethylene",
                                                            "Polystyrene","Standard"))

# Run ANCOMBC2 with standard as the reference
ancombc.starvation <- ancombc2(data = ps.rel.filtered,
                             fix_formula = "Treatment", 
                             rand_formula = "(1 | Insect_batch/Replicate)",
                             p_adj_method = "BH",
                             alpha = 0.025,
                             global = FALSE)
ancombc.starvation.res <- ancombc.starvation$res

# Get differentialy abundant taxa for PE
taxa.diff_abn.PE = data.frame()
for(n in 1:nrow(ancombc.standard.res)) { 
  if(ancombc.standard.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.standard.res$passed_ss_TreatmentPolyethylene[n] == TRUE &
     ancombc.starvation.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.starvation.res$passed_ss_TreatmentPolyethylene[n] == TRUE) { 
    taxa.diff_abn.PE <- rbind(taxa.diff_abn.PE, res.asv$taxon[n])
  }
}

# Print the results for PE
if(nrow(taxa.diff_abn.PE) == 0) {
  print("No MAGs met the criteria for differential abundance for PE")
} else {
  print(taxa.diff_abn.PE)
}

# Get differentialy abundant taxa for PS
taxa.diff_abn.PS = data.frame()
for(n in 1:nrow(ancombc.standard.res)) { 
  if(ancombc.standard.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.standard.res$passed_ss_TreatmentPolyethylene[n] == TRUE &
     ancombc.starvation.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.starvation.res$passed_ss_TreatmentPolyethylene[n] == TRUE) { 
    taxa.diff_abn.PS <- rbind(taxa.diff_abn.PS, res.asv$taxon[n])
  }
}

# Print the results for PS
if(nrow(taxa.diff_abn.PS) == 0) {
  print("No MAGs met the criteria for differential abundance for PS")
} else {
  print(taxa.diff_abn.PS)
}

```

Summarize the data at the species level and then test whether any species are enriched in the PE or PS diets relative to both the starvation and standard diets.

```{r Mealworm species differential abundance, message=FALSE, warning=FALSE}

# Summarize at the species level, removing taxa not classified at the species level
ps.rel.species <- tax_glom(ps.rel, "Species", NArm = TRUE)

# Set abundance filter function, filtering based on an average relative abundance ≥ 0.01%
abundance_filter_function <- function(x) {
  mean(x) >= 0.0001
}

# Filter the MAG data to a minimum average relative abundance of 0.01%
ps.rel.filtered <- ps

# Set the factors to get standard as the reference
sample_data(ps.rel.filtered)$Treatment <- factor(sample_data(ps.rel.filtered)$Treatment, 
                                                 levels = c("Standard", "Polyethylene",
                                                            "Polystyrene","Starvation"))

# Run ANCOMBC2 with standard as teh reference
ancombc.standard <- ancombc2(data = ps.rel.filtered,
                             fix_formula = "Treatment", 
                             rand_formula = "(1 | Insect_batch/Replicate)",
                             p_adj_method = "BH",
                             alpha = 0.025,
                             global = FALSE)
ancombc.standard.res <- ancombc.standard$res

# Set the factors to get standard as the reference
sample_data(ps.rel.filtered)$Treatment <- factor(sample_data(ps.rel.filtered)$Treatment, 
                                                 levels = c("Starvation", "Polyethylene",
                                                            "Polystyrene","Standard"))

# Run ANCOMBC2 with standard as the reference
ancombc.starvation <- ancombc2(data = ps.rel.filtered,
                             fix_formula = "Treatment", 
                             rand_formula = "(1 | Insect_batch/Replicate)",
                             p_adj_method = "BH",
                             alpha = 0.025,
                             global = FALSE)
ancombc.starvation.res <- ancombc.starvation$res

# Get differentialy abundant taxa for PE
taxa.diff_abn.species.PE = data.frame()
for(n in 1:nrow(ancombc.standard.res)) { 
  if(ancombc.standard.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.standard.res$passed_ss_TreatmentPolyethylene[n] == TRUE &
     ancombc.starvation.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.starvation.res$passed_ss_TreatmentPolyethylene[n] == TRUE) { 
    taxa.diff_abn.species.PE <- rbind(taxa.diff_abn.species.PE, res.asv$taxon[n])
  }
}

# Print the results for PE
if(nrow(taxa.diff_abn.species.PE) == 0) {
  print("No species met the criteria for differential abundance for PE")
} else {
  print(taxa.diff_abn.species.PE)
}

# Get differentialy abundant taxa for PS
taxa.diff_abn.species.PS = data.frame()
for(n in 1:nrow(ancombc.standard.res)) { 
  if(ancombc.standard.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.standard.res$passed_ss_TreatmentPolyethylene[n] == TRUE &
     ancombc.starvation.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.starvation.res$passed_ss_TreatmentPolyethylene[n] == TRUE) { 
    taxa.diff_abn.species.PS <- rbind(taxa.diff_abn.species.PS, res.asv$taxon[n])
  }
}

# Print the results for PS
if(nrow(taxa.diff_abn.species.PS) == 0) {
  print("No species met the criteria for differential abundance for PS")
} else {
  print(taxa.diff_abn.species.PS)
}

```

## Clean the workspace to switch to superworm data

```{r Clear mealworm data, results='hide', message=FALSE, warning=FALSE}

rm(list = ls())

```



# Analyse of the superworm data across treatments
## Prepare a phyloseq object for the superworm data

Use the MAG abundance data and the GTDB-Tk output data to generate a phyloseq object.

```{r Superworm phyloseq object, results='hide', message=FALSE, warning=FALSE}

# Load the OTU table
otumat <- as.matrix(read.table("superworm_MAG_abundances_depth.txt",
                        header = TRUE, row.names = 1))

# Load the taxonomy table
taxmat <- as.matrix(read.table("superworm_MAG_taxonomy.txt",
                        header = TRUE, row.names = 1))

# Load the sample metadata
metadata <- read.table("superworm_sample_metadata.txt",
                       header = TRUE, row.names = 1)

# Combine the data into a phyloseq object
OTU <- otu_table(otumat, taxa_are_rows = TRUE)
TAX <- tax_table(taxmat)
META <- sample_data(metadata)
ps <- phyloseq(OTU, TAX, META)

# Convert MAG abundances to percentages
ps.rel <- transform_sample_counts(ps, function(x) x / sum(x))

```

## Calculate alpha-diversity for the superworm data and perform statistical analysis

Calculate Shannon and Simpson's diversity metrics and test if the values are statistically different between samples. Include insect batch and replicate as random effects, with replicate nested within batch.

```{r Superworm alpha diversity stats, message=FALSE, warning=FALSE}


# Calculate alpha-diversity metrics
richness <- estimate_richness(ps.rel, measures = c("Shannon", "Simpson"))

# Add columns with relevant metadata
richness$treatment <- metadata$Treatment
richness$replicate <- metadata$Replicate
richness$batch <- metadata$Insect_batch

# Create factors for treatment, ordering the treatments as desired
richness$treatment <- factor(richness$treatment, 
                                 levels = c("Standard", "Polyethylene", "Polystyrene", "Starvation"))

# Perform statistical analysis on the Shannon data
model.shannon <- lmer(log(Shannon) ~ treatment + (1 | batch/replicate), data=richness)
summary(model.shannon)
model.shannon.aov <- Anova(model.shannon, type = 2)
model.shannon.em <- emmeans(model.shannon, pairwise ~ treatment,
                            infer = TRUE, type = "response")
model.shannon.em$emmeans
model.shannon.em$contrasts

# Perform statistical analysis on the Simpson data
model.simpson <- lmer(log(Simpson) ~ treatment + (1 | batch/replicate), data=richness)
summary(model.simpson)
model.simpson.aov <- Anova(model.simpson, type = 2)
model.simpson.em <- emmeans(model.simpson, pairwise ~ treatment,
                            infer = TRUE, type = "response")
model.simpson.em$emmeans
model.simpson.em$contrasts

```

## Plot the alpha-diversity metrics for the superworm samples

Generate plots showing the Shannon and Simpson's diversity metrics

```{r Superworm alpha diversity plots, results='hide', message=FALSE, warning=FALSE}

# Collect the emmeans for Shannon data
emms.shannon <- as.data.frame(model.shannon.em$emmeans)
emms.shannon$metric <- 'Shannon'

# Collect the emmeans for teh Simpson's data
emms.simpson <- as.data.frame(model.simpson.em$emmeans)
emms.simpson$metric <- 'Simpson'

# Plot the Shannon data
p.shannon <- ggplot() +
  geom_bar(data = emms.shannon,
           aes(x = treatment, y = response, fill = treatment),
           stat = "identity",
           colour = "black") +
  geom_errorbar(data = emms.shannon,
                aes(x = treatment, ymin = response-SE, ymax = response+SE),
                stat = "identity", width = 0.5, size = 0.5) +
  geom_jitter(data = richness,
              aes(x = treatment, y = Shannon),
              width = 0.15, size = 1.25) +
  facet_wrap(~metric, scales="free_y") +
  theme_bw() +
  theme(text = element_text(size = 17), legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(face="bold"),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1, color="black"),
        axis.text.y = element_text(color="black")) +
  scale_y_continuous(lim=c(0,3.5), breaks=c(0.0,0.7,1.4,2.1,2.8,3.5)) + 
  xlab("") + ylab("")

# Plot the Simpson's data
p.simpson <- ggplot() +
  geom_bar(data = emms.simpson,
           aes(x = treatment, y = response, fill = treatment),
           stat = "identity",
           colour = "black") +
  geom_errorbar(data = emms.simpson,
                aes(x = treatment, ymin = response-SE, ymax = response+SE),
                stat = "identity", width = 0.5, size = 0.5) +
  geom_jitter(data = richness,
              aes(x = treatment, y = Simpson),
              width = 0.15, size = 1.25) +
  facet_wrap(~metric, scales="free_y") +
  theme_bw() +
  theme(text = element_text(size = 17), legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(face="bold"),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1, color="black"),
        axis.text.y = element_text(color="black")) +
  scale_y_continuous(lim=c(0,1), breaks=c(0.0,0.2,0.4,0.6,0.8,1.0)) + 
  xlab("") + ylab("")

# Combine the plots as two panels in one figure
p.shannon + p.simpson

# Export a PDF of the figure
pdf('Superworm.alpha.pdf', width=10, height=5)
p.shannon + p.simpson
dev.off()

```

## Calculate and plot beta-diversity for the superworm data and perform statistical analysis

```{r Superworm beta diversity permdisp, message=FALSE, warning=FALSE}

# Calculate the bray curtis distances
dist_bc <- phyloseq::distance(ps.rel, method = "bray")

# Extract sample data
meta <- data.frame(sample_data(ps.rel))

# Perform a permutation test to test if the data meet the assumptions of PERMANOVA
beta.bray <- betadisper(dist_bc, meta$Treatment, type = "centroid", bias.adjust = TRUE)
permdisp <- permutest(beta.bray, pairwise = TRUE, permutations = 10000)
permdisp

```

Unlike the mealworm data, the dispersion is not significantly different between samples, which means the results of PERMANOVA should be reliable.

```{r Superworm beta diversity permanova, message=FALSE, warning=FALSE}

# Run the PERMANOVA test
adonis.bray <- adonis2(dist_bc ~ Treatment + Insect_batch, data = meta, permutations = 10000, by = "terms")
as.data.frame(adonis.bray)

```

```{r Superworm beta diversity figure, results='hide', message=FALSE, warning=FALSE}

# Create a PCoA plot
sample_data(ps.rel)$Treatment <- factor(sample_data(ps.rel)$Treatment, 
                                        levels = c("Standard", "Polyethylene", "Polystyrene", "Starvation"))
pso <- ordinate(ps.rel, "PCoA", "bray")
var_explained <- pso$values$Relative_eig * 100
ordplot5 <- plot_ordination(ps.rel, pso, type="samples", 
                            color="Treatment", shape="Insect_batch",
                            axes = c(1,2))
p.beta.pcoa <- ordplot5 +  theme_bw() + theme(text = element_text(size = 17), aspect.ratio = 1) + 
  geom_point(size = 5) +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right") +
  labs(x = paste0(" PC1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" PC2 [", round(var_explained[2], 1), "%]"))
p.beta.pcoa

# Save the figure
pdf('Superworm.beta.pcoa.pdf', width=10, height=5)
p.beta.pcoa
dev.off()

```

Now run a capscale analysis since it is less sensitive to differences in dispersion and thus should be more appropriate than PERMANOVA in this case.

```{r Superworm beta diversity capscale, message=FALSE, warning=FALSE}

# Set the formula for the model
class(fo <- dist_bc ~ meta$Treatment + meta$Insect_batch)

# Run the capscale analysis followed by an ANOVA
pso_cap <- ordinate(ps.rel, "CAP", "bray", formula=fo)
anova(pso_cap, by = "terms", type = 2)

```

```{r Superworm beta diversity capscale figure, results='hide', message=FALSE, warning=FALSE}

# Get the relative eigenvalues
all_eig <- eigenvals(pso_cap)
rel_eig <- all_eig / sum(all_eig)
var_explained <- round(rel_eig * 100, 1)

# Plot the capscale results
ordplot5 <- plot_ordination(ps.rel, pso_cap, type="samples", color="Treatment",
                            shape="Insect_batch", axes = c(1,2))
p.beta.cap <- ordplot5 +  theme_bw() + theme(text = element_text(size = 17), aspect.ratio = 1) + 
  geom_point(size = 5) +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right") +
  labs(x = paste0(" CAP 1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" CAP 2 [", round(var_explained[2], 1), "%]"))
p.beta.cap

# Save the figure
pdf('Superworm.beta.cap.pdf', width=10, height=5)
p.beta.pcoa
dev.off()

```

Given the strong impact of batch, see what the PCoA and CAP plots look like if summarized at the genus level

```{r Superworm beta diversity genus, message=FALSE, warning=FALSE}

# Summarize at the genus level, removing taxa not classified at the genus level
ps.rel.genus <- tax_glom(ps.rel, "Genus", NArm = TRUE)

# Calculate the bray curtis distances
dist_bc <- phyloseq::distance(ps.rel.genus, method = "bray")

# Extract sample data
meta <- data.frame(sample_data(ps.rel.genus))

# Perform a permutation test to test if the data meet the assumptions of PERMANOVA
beta.bray <- betadisper(dist_bc, meta$Treatment, type = "centroid", bias.adjust = TRUE)
permdisp <- permutest(beta.bray, pairwise = TRUE, permutations = 10000)
permdisp

# Run the PERMANOVA test
adonis.bray <- adonis2(dist_bc ~ Treatment + Insect_batch, data = meta, permutations = 10000, by = "terms")
as.data.frame(adonis.bray)

# Create a PCoA plot
pso <- ordinate(ps.rel.genus, "PCoA", "bray")
var_explained <- pso$values$Relative_eig * 100
ordplot5 <- plot_ordination(ps.rel.genus, pso, type="samples", 
                            color="Treatment", shape="Insect_batch", 
                            axes = c(1,2))
p.beta.pcoa.genus <- ordplot5 + theme_bw() + 
  theme(text = element_text(size = 17), aspect.ratio = 1) +
  geom_point(size = 5) +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = paste0(" PC1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" PC2 [", round(var_explained[2], 1), "%]"))

# Set the formula for the model for the capscale analysis
class(fo <- dist_bc ~ meta$Treatment + meta$Insect_batch)

# Run the capscale analysis followed by an ANOVA
pso_cap <- ordinate(ps.rel.genus, "CAP", "bray", formula=fo)
anova(pso_cap, by = "terms", type = 2)

# Get the relative eigenvalues
all_eig <- eigenvals(pso_cap)
rel_eig <- all_eig / sum(all_eig)
var_explained <- round(rel_eig * 100, 1)

# Plot the capscale results
ordplot5 <- plot_ordination(ps.rel.genus, pso_cap, type="samples", color="Treatment",
                            shape="Insect_batch", axes = c(1,2))
p.beta.cap.genus <- ordplot5 + theme_bw() + 
  theme(text = element_text(size = 17), aspect.ratio = 1) +
  geom_point(size = 5) +
  theme_bw() +
  theme(text = element_text(size = 17),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = paste0(" CAP 1 [", round(var_explained[1], 1), "%]"),
       y = paste0(" CAP 2 [", round(var_explained[2], 1), "%]"))

```

```{r Superworm beta diversity genus figure, results='hide', message=FALSE, warning=FALSE}

# Display both graphs
p.beta.pcoa.genus + p.beta.cap.genus

# Save the figure
pdf('Superworm.beta.genus.pdf', width=10, height=5)
p.beta.pcoa.genus + p.beta.cap.genus
dev.off()

```

## Compare taxon abundance in the superworm data

Summarize the most abundant taxa at the phylum level in the superworms across diets.

```{r Superworm phylum abundance, message=FALSE, warning=FALSE}

# Agglomerate data at phylum level 
ps.rel.phylum <- tax_glom(ps.rel, "Phylum", NArm = FALSE)

# Create a dataframe from phyloseq object
ps.rel.phylum.df <- psmelt(ps.rel.phylum)

# Get summary statistics
df <- ps.rel.phylum.df %>%
  group_by(Treatment, Phylum) %>%
  summarize(mean = mean(Abundance, na.rm = TRUE),
            median = median(Abundance, na.rm = TRUE),
            stdev = sd(Abundance, na.rm = TRUE),
            counts = n(),
            se = stdev / sqrt(counts))

# Create a table rounded to 1 decimal place
df.w <- df %>%
  mutate(stat = paste0(round(100 * mean, 1), " +/- ", round(100 * stdev, 1))) %>%
  select(Treatment, Phylum, stat) %>%
  pivot_wider(
    names_from = Treatment,
    values_from = stat
  )
kable(df.w, digits = 1, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

# Create a table rounded to 6 decimal place
df.w <- df %>%
  mutate(stat = paste0(round(100 * mean, 6), " +/- ", round(100 * stdev, 6))) %>%
  select(Treatment, Phylum, stat) %>%
  pivot_wider(
    names_from = Treatment,
    values_from = stat
  )
kable(df.w, digits = 6, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

# Get overall summary statistics not divided by diet
df <- ps.rel.phylum.df %>%
  group_by(Phylum) %>%
  summarize(mean = 100 * mean(Abundance, na.rm = TRUE),
            stdev = 100 * sd(Abundance, na.rm = TRUE),
            counts = n(),
            se = stdev / sqrt(counts))
kable(df, digits = 6, align = "l",
      caption = "test") %>%
  kable_classic(full_width = FALSE)

```

## Differential abundance analysis of taxa in the superworm data

```{r Superworm MAG differential abundance, message=FALSE, warning=FALSE}

# Set abundance filter function, filtering based on an average relative abundance ≥ 0.01%
abundance_filter_function <- function(x) {
  mean(x) >= 0.0001
}

# Filter the MAG data to a minimum average relative abundance of 0.01%
ps.rel.filtered <- filter_taxa(ps.rel, abundance_filter_function, prune = TRUE)

# Set the factors to get standard as the reference
sample_data(ps.rel.filtered)$Treatment <- factor(sample_data(ps.rel.filtered)$Treatment, 
                                                 levels = c("Standard", "Polyethylene",
                                                            "Polystyrene","Starvation"))

# Run ANCOMBC2 with standard as teh reference
ancombc.standard <- ancombc2(data = ps.rel.filtered,
                             fix_formula = "Treatment", 
                             rand_formula = "(1 | Insect_batch/Replicate)",
                             p_adj_method = "BH",
                             alpha = 0.025,
                             global = FALSE)
ancombc.standard.res <- ancombc.standard$res

# Set the factors to get standard as the reference
sample_data(ps.rel.filtered)$Treatment <- factor(sample_data(ps.rel.filtered)$Treatment, 
                                                 levels = c("Starvation", "Polyethylene",
                                                            "Polystyrene","Standard"))

# Run ANCOMBC2 with standard as the reference
ancombc.starvation <- ancombc2(data = ps.rel.filtered,
                             fix_formula = "Treatment", 
                             rand_formula = "(1 | Insect_batch/Replicate)",
                             p_adj_method = "BH",
                             alpha = 0.025,
                             global = FALSE)
ancombc.starvation.res <- ancombc.starvation$res

# Get differentialy abundant taxa for PE
taxa.diff_abn.PE = data.frame()
for(n in 1:nrow(ancombc.standard.res)) { 
  if(ancombc.standard.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.standard.res$passed_ss_TreatmentPolyethylene[n] == TRUE &
     ancombc.starvation.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.starvation.res$passed_ss_TreatmentPolyethylene[n] == TRUE) { 
    taxa.diff_abn.PE <- rbind(taxa.diff_abn.PE, res.asv$taxon[n])
  }
}

# Print the results for PE
if(nrow(taxa.diff_abn.PE) == 0) {
  print("No MAGs met the criteria for differential abundance for PE")
} else {
  print(taxa.diff_abn.PE)
}

# Get differentialy abundant taxa for PS
taxa.diff_abn.PS = data.frame()
for(n in 1:nrow(ancombc.standard.res)) { 
  if(ancombc.standard.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.standard.res$passed_ss_TreatmentPolyethylene[n] == TRUE &
     ancombc.starvation.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.starvation.res$passed_ss_TreatmentPolyethylene[n] == TRUE) { 
    taxa.diff_abn.PS <- rbind(taxa.diff_abn.PS, res.asv$taxon[n])
  }
}

# Print the results for PS
if(nrow(taxa.diff_abn.PS) == 0) {
  print("No MAGs met the criteria for differential abundance for PS")
} else {
  print(taxa.diff_abn.PS)
}

```

Summarize the data at the species level and then test whether any species are enriched in the PE or PS diets relative to both the starvation and standard diets.

```{r Superworm species differential abundance, message=FALSE, warning=FALSE}

# Summarize at the species level, removing taxa not classified at the species level
ps.rel.species <- tax_glom(ps.rel, "Species", NArm = TRUE)

# Set abundance filter function, filtering based on an average relative abundance ≥ 0.01%
abundance_filter_function <- function(x) {
  mean(x) >= 0.0001
}

# Filter the MAG data to a minimum average relative abundance of 0.01%
ps.rel.filtered <- filter_taxa(ps.rel.species, abundance_filter_function, prune = TRUE)

# Set the factors to get standard as the reference
sample_data(ps.rel.filtered)$Treatment <- factor(sample_data(ps.rel.filtered)$Treatment, 
                                                 levels = c("Standard", "Polyethylene",
                                                            "Polystyrene","Starvation"))

# Run ANCOMBC2 with standard as teh reference
ancombc.standard <- ancombc2(data = ps.rel.filtered,
                             fix_formula = "Treatment", 
                             rand_formula = "(1 | Insect_batch/Replicate)",
                             p_adj_method = "BH",
                             alpha = 0.025,
                             global = FALSE)
ancombc.standard.res <- ancombc.standard$res

# Set the factors to get standard as the reference
sample_data(ps.rel.filtered)$Treatment <- factor(sample_data(ps.rel.filtered)$Treatment, 
                                                 levels = c("Starvation", "Polyethylene",
                                                            "Polystyrene","Standard"))

# Run ANCOMBC2 with standard as the reference
ancombc.starvation <- ancombc2(data = ps.rel.filtered,
                             fix_formula = "Treatment", 
                             rand_formula = "(1 | Insect_batch/Replicate)",
                             p_adj_method = "BH",
                             alpha = 0.025,
                             global = FALSE)
ancombc.starvation.res <- ancombc.starvation$res

# Get differentialy abundant taxa for PE
taxa.diff_abn.species.PE = data.frame()
for(n in 1:nrow(ancombc.standard.res)) { 
  if(ancombc.standard.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.standard.res$passed_ss_TreatmentPolyethylene[n] == TRUE &
     ancombc.starvation.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.starvation.res$passed_ss_TreatmentPolyethylene[n] == TRUE) { 
    taxa.diff_abn.species.PE <- rbind(taxa.diff_abn.species.PE, res.asv$taxon[n])
  }
}

# Print the results for PE
if(nrow(taxa.diff_abn.species.PE) == 0) {
  print("No species met the criteria for differential abundance for PE")
} else {
  print(taxa.diff_abn.species.PE)
}

# Get differentialy abundant taxa for PS
taxa.diff_abn.species.PS = data.frame()
for(n in 1:nrow(ancombc.standard.res)) { 
  if(ancombc.standard.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.standard.res$passed_ss_TreatmentPolyethylene[n] == TRUE &
     ancombc.starvation.res$diff_TreatmentPolyethylene[n] == TRUE & 
     ancombc.starvation.res$passed_ss_TreatmentPolyethylene[n] == TRUE) { 
    taxa.diff_abn.species.PS <- rbind(taxa.diff_abn.species.PS, res.asv$taxon[n])
  }
}

# Print the results for PS
if(nrow(taxa.diff_abn.species.PS) == 0) {
  print("No species met the criteria for differential abundance for PS")
} else {
  print(taxa.diff_abn.species.PS)
}

```

## Clean the workspace to switch to CFU analysis

```{r Clear superworm data, results='hide', message=FALSE, warning=FALSE}

rm(list = ls())

```

# Analysis of CFU counts per insect
